---
title: "Library Management"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Library Management}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The **libr** package also contains a number of functions for managing
data libraries.  There are functions to add and remove data from a library,
as well as copy or delete an entire library.

The example below illustrates some of these management functions.  The example
first creates a library and adds some data to it.  The example then proceeds
to copy it to another library and perform some manipulation of the data
in the libraries.  The example ends by looking at some of the metadata 
available for libraries:

```{r eval=FALSE, echo=TRUE}
library(libr)

# Create temp directory
tmp <- tempdir()

# Create libraries
libname(s1, tmp)

# Add data to library and adjust names
lib_add(s1, state.name, state.area, state.region, state.abb,
        name = c("name", "area", "region", "abb"))
# # library 's1': 4 items
# - attributes: rds not loaded
# - path: C:\Users\User\AppData\Local\Temp\RtmpqAMV6L
# - items:
#     Name Extension Rows Cols   Size        LastModified
# 1   name       rds   50    1 4.4 Kb 2020-11-29 17:00:28
# 2   area       rds   50    1 1.4 Kb 2020-11-29 17:00:28
# 3 region       rds   50    1 1.9 Kb 2020-11-29 17:00:28
# 4    abb       rds   50    1 4.1 Kb 2020-11-29 17:00:28

# Copy library to backup location
lib_copy(s1, s2, file.path(tmp, "orig"))
# # library 's2': 4 items
# - attributes: rds not loaded
# - path: C:\Users\User\AppData\Local\Temp\RtmpqAMV6L/orig
# - items:
#     Name Extension Rows Cols   Size        LastModified
# 1   name       rds   50    1 4.4 Kb 2020-11-29 17:01:17
# 2   area       rds   50    1 1.4 Kb 2020-11-29 17:01:17
# 3 region       rds   50    1 1.9 Kb 2020-11-29 17:01:17
# 4    abb       rds   50    1 4.1 Kb 2020-11-29 17:01:17

# Remove data from library 1
lib_remove(s1, name = c("name", "area", "region", "abb"))
# # library 's1': 0 items
# - attributes: rds not loaded
# - path: C:\Users\User\AppData\Local\Temp\RtmpqAMV6L
# NULL

# Load library 1 into memory
lib_load(s1)

s1.combined <- data.frame(name = s2.name, abb = s2.abb, 
                              area = s2.area, region = s2.region)

s1.east <- subset(s1.combined, region == "Northeast")
s1.west <- subset(s1.combined, region == "West")
s1.north <- subset(s1.combined, region == "North Central")
s1.south <- subset(s1.combined, region == "South")

# Sync workspace with library list
lib_sync(s1)
# # library 's1': 5 items
# - attributes: rds loaded
# - path: C:\Users\User\AppData\Local\Temp\RtmpqAMV6L
# - items:
#       Name Extension Rows Cols    Size LastModified
# 1 combined        NA   50    4 12.7 Kb         <NA>
# 2     east        NA    9    4  3.2 Kb         <NA>
# 3    north        NA   12    4  3.5 Kb         <NA>
# 4    south        NA   16    4    4 Kb         <NA>
# 5     west        NA   13    4  3.6 Kb         <NA>

# Save library 1 to disk
lib_write(s1)
# # library 's1': 5 items
# - attributes: rds loaded
# - path: C:\Users\User\AppData\Local\Temp\RtmpqAMV6L
# - items:
#       Name Extension Rows Cols    Size        LastModified
# 1 combined       rds   50    4 13.4 Kb 2020-11-29 17:03:40
# 2     east       rds    9    4    4 Kb 2020-11-29 17:03:40
# 3    north       rds   12    4  4.3 Kb 2020-11-29 17:03:40
# 4    south       rds   16    4  4.8 Kb 2020-11-29 17:03:40
# 5     west       rds   13    4  4.4 Kb 2020-11-29 17:03:40

# View path
lib_path(s1)
# [1] "C:\\Users\\User\\AppData\\Local\\Temp\\RtmpqAMV6L"

# View size
lib_size(s1)
# [1] 7175

# View info
lib_info(s1)
#       Name Extension Rows Cols    Size        LastModified
# 1 combined       rds   50    4 13.4 Kb 2020-11-29 17:03:40
# 2     east       rds    9    4    4 Kb 2020-11-29 17:03:40
# 3    north       rds   12    4  4.3 Kb 2020-11-29 17:03:40
# 4    south       rds   16    4  4.8 Kb 2020-11-29 17:03:40
# 5     west       rds   13    4  4.4 Kb 2020-11-29 17:03:40

# Display dictionary
dictionary(s1)
# # A tibble: 20 x 10
#    Name     Column Class     Label Description Format Width Justify  Rows   NAs
#    <chr>    <chr>  <chr>     <chr> <chr>       <lgl>  <int> <chr>   <int> <int>
#  1 combined name   character NA    NA          NA        14 NA         50     0
#  2 combined abb    character NA    NA          NA         2 NA         50     0
#  3 combined area   numeric   NA    NA          NA        NA NA         50     0
#  4 combined region factor    NA    NA          NA        NA NA         50     0
#  5 east     name   character NA    NA          NA        13 NA          9     0
#  6 east     abb    character NA    NA          NA         2 NA          9     0
#  7 east     area   numeric   NA    NA          NA        NA NA          9     0
#  8 east     region factor    NA    NA          NA        NA NA          9     0
#  9 north    name   character NA    NA          NA        12 NA         12     0
# 10 north    abb    character NA    NA          NA         2 NA         12     0
# 11 north    area   numeric   NA    NA          NA        NA NA         12     0
# 12 north    region factor    NA    NA          NA        NA NA         12     0
# 13 south    name   character NA    NA          NA        14 NA         16     0
# 14 south    abb    character NA    NA          NA         2 NA         16     0
# 15 south    area   numeric   NA    NA          NA        NA NA         16     0
# 16 south    region factor    NA    NA          NA        NA NA         16     0
# 17 west     name   character NA    NA          NA        10 NA         13     0
# 18 west     abb    character NA    NA          NA         2 NA         13     0
# 19 west     area   numeric   NA    NA          NA        NA NA         13     0
# 20 west     region factor    NA    NA          NA        NA NA         13     0

# Unload library 1
lib_unload(s1)

# Clean up
lib_delete(s1)
lib_delete(s2)
```


Next: [Data Step Operations](https://libr.r-sassy.org/articles/libr-datastep.html)
