---
title: "Data Step Operations"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Data Step Operations}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
Normally, R processes data column-by-column. The data step allows you 
to process data row-by-row.  Row-by-row processing of data is useful when you 
have related columns, and wish to perform conditional logic on those 
columns. The `datastep()` function allows you to realize this style of 
data processing. It is particularly advantageous when you wish to perform deeply 
nested conditional logic. It is also very useful for by-group
processing.

#### Example 1: Simple Data Step
Here is an example of a simple data step:
```{r eval=FALSE, echo=TRUE}
library(libr)

# Add some columns to mtcars using data step logic
df <- datastep(mtcars[1:10, 1:3], {

    if (mpg >= 20) 
      mpgcat <- "High"
    else 
      mpgcat <- "Low"
      
    recdt <- as.Date("1974-06-10")
    
    if (cyl == 8)
      is8cyl <- TRUE

  })
  
# View results  
df
#                    mpg cyl  disp      recdt mpgcat is8cyl
# Mazda RX4         21.0   6 160.0 1974-06-10   High     NA
# Mazda RX4 Wag     21.0   6 160.0 1974-06-10   High     NA
# Datsun 710        22.8   4 108.0 1974-06-10   High     NA
# Hornet 4 Drive    21.4   6 258.0 1974-06-10   High     NA
# Hornet Sportabout 18.7   8 360.0 1974-06-10    Low   TRUE
# Valiant           18.1   6 225.0 1974-06-10    Low     NA
# Duster 360        14.3   8 360.0 1974-06-10    Low   TRUE
# Merc 240D         24.4   4 146.7 1974-06-10   High     NA
# Merc 230          22.8   4 140.8 1974-06-10   High     NA
# Merc 280          19.2   6 167.6 1974-06-10    Low     NA

```
### Keep, Drop, and Rename 

The data step has parameters to perform basic shaping of the resulting
data frame.  These parameters are 'keep', 'drop', and 'rename'.  For example,
the above data step could have been performed by sending all columns into 
the data step, and keeping only the desired columns.  Using the `keep` 
parameter also allows you to order the resulting columns.

#### Example 2: Keeping Data Step Variables
```{r eval=FALSE, echo=TRUE}
library(libr)

# Keep and order output columns 
df <- datastep(mtcars[1:10,], 
  keep = c("mpg", "cyl", "disp", "mpgcat", "recdt"), {

    if (mpg >= 20) 
      mpgcat <- "High"
    else 
      mpgcat <- "Low"
      
    recdt <- as.Date("1974-06-10")
    
    if (cyl == 8)
      is8cyl <- TRUE

  })
  
df
#                    mpg cyl  disp mpgcat      recdt
# Mazda RX4         21.0   6 160.0   High 1974-06-10
# Mazda RX4 Wag     21.0   6 160.0   High 1974-06-10
# Datsun 710        22.8   4 108.0   High 1974-06-10
# Hornet 4 Drive    21.4   6 258.0   High 1974-06-10
# Hornet Sportabout 18.7   8 360.0    Low 1974-06-10
# Valiant           18.1   6 225.0    Low 1974-06-10
# Duster 360        14.3   8 360.0    Low 1974-06-10
# Merc 240D         24.4   4 146.7   High 1974-06-10
# Merc 230          22.8   4 140.8   High 1974-06-10
# Merc 280          19.2   6 167.6    Low 1974-06-10

```
### The Retain Parameter

The retain parameter allows you to define variables
that will be seeded with the value from the previous step. The retain
option is useful for creating cumulative values or for performing 
conditions based on the value of the previous row.

#### Example 3: Drop, Retain, and Rename Parameters
```{r eval=FALSE, echo=TRUE}
library(libr)

df <- datastep(mtcars[1:10, ],
               drop = c("disp", "hp", "drat", "qsec",
                        "vs", "am", "gear", "carb"),
               retain = list(cumwt = 0 ),
               rename = c(mpg = "MPG", cyl = "Cylinders", wt = "Wgt",
                          cumwt = "Cumulative Wgt"), {

  cumwt <- cumwt + wt

 })

df
#                    MPG Cylinders   Wgt Cumulative Wgt
# Mazda RX4         21.0         6 2.620          2.620
# Mazda RX4 Wag     21.0         6 2.875          5.495
# Datsun 710        22.8         4 2.320          7.815
# Hornet 4 Drive    21.4         6 3.215         11.030
# Hornet Sportabout 18.7         8 3.440         14.470
# Valiant           18.1         6 3.460         17.930
# Duster 360        14.3         8 3.570         21.500
# Merc 240D         24.4         4 3.190         24.690
# Merc 230          22.8         4 3.150         27.840
# Merc 280          19.2         6 3.440         31.280

```
### By Group Processing
The `datastep()` function also has the capabilities of performing by-group
processing.  A by-group is accomplished using the `by` parameter, and passing
a vector of column names that define the group.  Once a by-group is 
defined, the `first.` and `last.` automatic variables become active, which
allow you to identify the boundaries between groups.  Note that, by default,
your data must be sorted properly before sending it into the data step. To 
turn the sort check off, set the `sort_check` parameter to FALSE.

#### Example 4: By Groups
```{r eval=FALSE, echo=TRUE}
library(libr)

# Identify start and end of by-groups
df <- datastep(mtcars[1:10,], 
  keep = c("mpg", "cyl", "gear", "grp"), 
  by = c("gear"), sort_check = FALSE, {

    if (first. & last.)
      grp <- "Start - End"
    else if (first.)
      grp <- "Start"
    else if (last.)
      grp <- "End"
    else 
      grp <- "-"

  })
  
df
#                    mpg cyl gear   grp
# Mazda RX4         21.0   6    4 Start
# Mazda RX4 Wag     21.0   6    4     -
# Datsun 710        22.8   4    4   End
# Hornet 4 Drive    21.4   6    3 Start
# Hornet Sportabout 18.7   8    3     -
# Valiant           18.1   6    3     -
# Duster 360        14.3   8    3   End
# Merc 240D         24.4   4    4 Start
# Merc 230          22.8   4    4     -
# Merc 280          19.2   6    4   End

```
### Using Summary Functions
There may be times when you want to combine row-by-row conditional 
processing with column-by-column vector operations.  For example, 
let's say you want to calculate a mean and then perform conditional 
processing on that mean.  This 
situation can be handled using the `calculate` parameter on the `datastep()`
function. The function will execute the `calculate` block first, add any
assigned variables to the data frame, and then execute the data step.  Below 
is an example of such a scenario:

#### Example 5: Calculate Block
```{r eval=FALSE, echo=TRUE}
library(libr)

# Categorize mpg as above or below the mean
df <- datastep(mtcars, 
  keep = c("mpg", "cyl", "mean_mpg", "mpgcat"), 
  calculate = { mean_mpg = mean(mpg) },
  {

    if (mpg >= mean_mpg)
      mpgcat <- "High"
    else 
      mpgcat <- "Low"

  })
  
df[1:10,]
#                    mpg cyl mean_mpg mpgcat
# Mazda RX4         21.0   6 20.09062   High
# Mazda RX4 Wag     21.0   6 20.09062   High
# Datsun 710        22.8   4 20.09062   High
# Hornet 4 Drive    21.4   6 20.09062   High
# Hornet Sportabout 18.7   8 20.09062    Low
# Valiant           18.1   6 20.09062    Low
# Duster 360        14.3   8 20.09062    Low
# Merc 240D         24.4   4 20.09062   High
# Merc 230          22.8   4 20.09062   High
# Merc 280          19.2   6 20.09062    Low

```
### Data Steps with `dplyr`
Note that the `datastep()` function is pipe-friendly, and can be combined
with **dplyr** functions in a data pipeline.  Also note that the `datastep()`
function will recognize any group attributes added by the `group_by()` 
function.  Therefore, within a **dplyr** pipeline, it is not necessary to 
use any `datastep` parameters.  The following example recreates the above 
data frame from Example 5, but with a **dplyr** pipeline.

#### Example 6: Data Pipeline
```{r eval=FALSE, echo=TRUE}
library(libr)
library(dplyr)
library(magrittr)

# Add datastep to dplyr pipeline
df <- mtcars %>% 
  select(mpg, cyl, gear) %>% 
  mutate(mean_mpg = mean(mpg)) %>% 
  datastep({

    if (mpg >= mean_mpg)
      mpgcat <- "High"
    else 
      mpgcat <- "Low"

  }) %>% 
  filter(row_number() <= 10)
  
df
#     mpg cyl gear mean_mpg mpgcat
# 1  21.0   6    4 20.09062   High
# 2  21.0   6    4 20.09062   High
# 3  22.8   4    4 20.09062   High
# 4  21.4   6    3 20.09062   High
# 5  18.7   8    3 20.09062    Low
# 6  18.1   6    3 20.09062    Low
# 7  14.3   8    3 20.09062    Low
# 8  24.4   4    4 20.09062   High
# 9  22.8   4    4 20.09062   High
# 10 19.2   6    4 20.09062    Low

```

Next: [Enhanced Equality](https://libr.r-sassy.org/articles/libr-eq.html)


